---
title: "R Programming and Bio-conductor Assignment Report"
author: "Mutesasira Edward"
date: "Date: 2023-10-30"
output:
  pdf_document:
    toc_depth: '3'
  html_document:
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
# Importing packages
library(tidyverse)
library(psych)
library(ggtern)
library(report)
#Setting working directory
getwd()
setwd("/home/eddie/R/msb7102bioconductorandrprogrammingassignment")
# Reading the metaFile.csv
meta <- read_csv("metaFile.csv")
# Importing AssignmentFile.csv file
aFile <- read.csv("AssignmentFile.csv")
# Reading statsFile.csv file
stats <- read_csv("statsFile.csv")
```

### Section A: Data Exploration 1
metaFile.csv file was read into R session and the following were the explorations. 

The number of records and columns.
```{r dim1, echo=FALSE, comment=NA}
# Returning the number of rows and columns
dim(meta)
```

The number of samples which were obtained for each level of treatment
```{r count1, echo=FALSE, comment=NA}
# Counting the number of samples which were obtained for each level of treatment
tab<-table(meta$Treatment)
data.frame(Level=names(tab),Sample=as.vector(tab))
```

The number of samples which were obtained for each participant
```{r count2, echo=FALSE, comment=NA}
# Counting the number of samples which were obtained for each participant
tab<-table(meta$Participant)
data.frame(Participant=names(tab),Sample=as.vector(tab))
```

Records for samples obtained at treatment levels 0,2 and 4
```{r extract1, echo=FALSE, comment=NA}
# Extracting records for samples obtained at treatment levels 0,2 and 4
meta %>% 
  select(sample_id, Participant, Treatment)  %>%
  filter(Treatment %in% c(0, 2, 4))
```

Extract records for samples obtained from participants A and C
```{r extract2, echo=FALSE, comment=NA}
# Extracting records for samples obtained from participants A and C
meta %>% 
  select(sample_id, Participant, Treatment)  %>%
  filter(Participant %in% c("A","C"))
```

The number of samples under each participant groups per treatment level.
```{r table, echo=FALSE, comment=NA}
# Counting the number of samples under each participant groups per treatment level.
table(meta$Participant, meta$Treatment)
```

### Section B: Data Exploration 2
AssignmentFile.csv file was imported into R session and the following were the explorations.

The average concentration for each of the concentrations (ie Conc1, Conc2 and Conc3)
```{r mean1, echo=FALSE, comment=NA}
# Computing the average concentration for each of the concentrations (ie Conc1, Conc2 and Conc3)
mn <- lapply(select(aFile, Conc1,Conc2,Conc3), mean) 
df <- data.frame(mn)
row.names(df)[1] <- "Mean"  
df
```

The average Concentration for each sample
```{r mean2, echo=FALSE, comment=NA}
# Computing the average Concentration for each sample 
 rowMeans(aFile[,2:4], na.rm = T)
```

Extended data-frame with a new variable ”Concmean” to contain the average Concentration values computed above
```{r mean3, echo=FALSE, comment=NA}
# Extending data-frame with a new variable ”Concmean” to contain the average Concentration (mean) values computed above
aFile$Concmean <-  rowMeans(aFile[,2:4], na.rm = T)
aFile
```

The basic statics (min, max, mean, median, quartiles, variance, standard deviation) for all the variables including Concmean
```{r statics, echo=FALSE, comment=NA}
# Obtaining the basic statics (min, max, mean, median, quartiles, variance, standard deviation)
bstat1<-summary(aFile[,2:5])
bstat1
sd1<-as.data.frame(sapply(aFile[,2:5], sd))
scol<- colnames(sd1)[1] <- "Standard_Deviation" 
sd1

var1<-as.data.frame(sapply(aFile[,2:5], var))
vcol<-colnames(var1)[1] <- "Variance" 
var1

```

### Section C : Data Manipulation and Graphics
The new data-frame after removing variables Conc1, Conc2 and Conc3 
```{r new, echo=FALSE, comment=NA}
# Removing variables Conc1, Conc2, Conc3
new_aFile <- data.frame(aFile[, -c(2:4)])
new_aFile
```

Merged information in concentration data-frame with metadata data-frame 
```{r merged, echo=FALSE, comment=NA}
# Merging the concentration with metadata data-frames
names(new_aFile) <- c("sample_id", "Concmean")
#merged <- merge(meta, new_aFile, by = "sample_id", all = TRUE) 
merged<-cbind(meta, new_aFile[,2])
colnames(merged)[colnames(merged) == "new_aFile[, 2]"] <- "Concmean"
merged
```

Size of the combined data-frame (rows and columns)
```{r size, echo=FALSE, comment=NA}
# Checking the size of combined data-frame
dim(merged)
```

Variable “participantsBC” with the extracted records for samples for participants B and C obtained at treatment levels 2,3 and 4.
```{r participantBC, echo=FALSE, comment=NA}
# extracting records for samples for participants B and C obtained at treatment levels 2,3 and 4.
participantsBC <- merged %>%
  filter(Participant %in% c("B", "C") & Treatment %in% c(2, 3, 4)) %>%
  select(sample_id, Participant, Treatment, Concmean)
participantsBC
```

Computed average Concentration for participants B and C under treatments 2, 3 and 4. 
```{r average, echo=FALSE, comment=NA, message=FALSE}
# Computing the average Concentration for participants B and C under treatments 2, 3 and 4.
average_con  <- participantsBC %>%
  group_by(Participant, Treatment) %>%
  summarize(Average = mean(Concmean, na.rm = TRUE)) %>% 
  rename(Average_Concetration = Average)
print.data.frame(average_con)
```

Changed treatment levels from 0,1,2,3,4 to “very-low”, “low”, “moderate”, “high”, “very-high” respectively on the combined dataset using for loop.
```{r newLevels, echo=FALSE, comment=NA}
new_level <- c("very-low", "low", "moderate", "high", "very-high")
# Loop through Treatment column to change the levels
for (i in 1:length(new_level)) {
  merged$Treatment[merged$Treatment == i - 1] <- new_level[i]
}
print.data.frame(merged, row.names = FALSE)
```

Changed treatment levels from 0,1,2,3,4 to ”very-low”, ”low”, ”moderate”, ”high”, ”very-high” respectively using the created function
```{r function1, echo=FALSE, comment=NA}
# Function to map numeric treatment levels to categories
convert_treatment <- function(level) {
  level_labels <- c("very-low", "low", "moderate", "high", "very-high")
  names(level_labels) <- 0:4  # Assign names corresponding to numeric levels
  return(ifelse(level %in% names(level_labels), level_labels[as.character(level)], 
                as.character(level)))
}
merged$Treatment <- sapply(merged$Treatment, convert_treatment)
# Print the updated dataset
print.data.frame(merged, row.names = FALSE)
```

Time taken for for-loop to execute 
```{r time1, echo=FALSE, comment=NA}
start_time_for <- system.time({
treatment_level <- c("very-low", "low", "moderate", "high", "very-high")
# Loop through Treatment column to change the levels
for (i in 1:length(treatment_level)) {
  merged$Treatment[merged$Treatment == i - 1] <- treatment_level[i]
}
})
end_time_for <- system.time({})  # Measure execution time
execution_time_for <- end_time_for - start_time_for  # Calculate execution time
execution_time_for
```

Time taken for sapply() function to execute 
```{r time2, echo=FALSE, comment=NA}
start_time_sapply <- system.time({
convert_treatment <- function(level) {
  level_labels <- c("very-low", "low", "moderate", "high", "very-high")
  names(level_labels) <- 0:4  # Assign names corresponding to numeric levels
  return(ifelse(level %in% names(level_labels), level_labels[as.character(level)], 
                as.character(level)))
}
merged$Treatment <- sapply(merged$Treatment, convert_treatment)
})
end_time_sapply <- system.time({})  # Measure execution time

execution_time_sapply <- end_time_sapply - start_time_sapply  # Calculate execution time
execution_time_sapply
```
`COMMENT:` system.time() measures CPU time used by a specific expression or function and the outputs are with variables user, system and elapsed. User Time is the wall clock time. The time that a user experienced. Elapsed Time is the time charged to the CPU(s) for the expression. If elapsed time > user time, this means that the CPU is waiting around for some other operations (may be external) to be done. If elapsed time < user time, this means that the machine has multiple cores and is able to use them. Comparing the above time, sapply() function is faster than for-loop function.


Graphical representations of the combined dataset with the changed treatment levels. 
```{r graph1, echo=FALSE, comment=NA, message=TRUE}
ggplot(data = merged) +
geom_bar(mapping = aes(x = Treatment, fill = Participant))+labs(title = "The Graph treatment levels Filled by Participant. ", x = "Treatment level")  + theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Title appearance
    panel.grid.major = element_line(color = "gray"),  # Adjusting gridlines
    axis.text = element_text(color = "black"),  # Adjusting axis text color
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  )

ggplot(data = merged) +
geom_bar(mapping = aes(x =Participant, fill = Treatment),position = "dodge")+labs(title ="The Graph Participant Filled by Treatment level ", x = "Participant") + theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Title appearance
    panel.grid.major = element_line(color = "gray"),  # Adjusting gridlines
    axis.text = element_text(color = "black"),  # Adjusting axis text color
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  )

ggplot(data = merged) +
geom_point(mapping = aes(x = Participant, y = Concmean, shape = Treatment))+labs(title ="The ggplot of Concmean against Participant", x = "Participant", t="Concmean")  + theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Title appearance
    panel.grid.major = element_line(color = "gray"),  # Adjusting gridlines
    axis.text = element_text(color = "black"),  # Adjusting axis text color
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  )

ggplot(data = merged, mapping = aes(x = Treatment, y = Concmean)) +
geom_boxplot()+labs(title ="The ggplot of Concmean against Treatement", x = "Treatment", t="Concmean")  + theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Title appearance
    panel.grid.major = element_line(color = "gray"),  # Adjusting gridlines
    axis.text = element_text(color = "black"),  # Adjusting axis text color
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  )
```

### Section D : Classical hypothesis testing and statistical models in R
The statsFile.csv file that contains information about participants
involved in a study was read into a data-frame to explore the key predictors of a disease status

The relation of BMI to BP for the all study participants
```{r relation1, echo=FALSE, comment=NA}
#Performing correlation on BMI and BP
cor.test(stats$BMI, stats$BP)
cor.test(stats$BMI, stats$BP, method = "pearson")
# Performing the covariance
cov(stats$BMI, stats$BP)
cov(stats$BMI, stats$BP) / (sd(stats$BMI) * sd(stats$BP))
##comparison of means
mean(stats$BMI)
mean(stats$BP)

t.test(stats$BMI, stats$BP)
##comparison of variances
var(stats$BMI)
var(stats$BP)
var.test(stats$BMI, stats$BP)
```

How disease status (phenotype) is related to gender
```{r relation2, echo=FALSE, comment=NA}
# Performing chi-squared contingency table tests and goodness-of-fit tests.
chisq.test(table(stats$Status, stats$Gender))
table(stats$Status, stats$Gender)#table of status by gender
```

The constructed model to show the relationship between BMI and BP
```{r model, echo=FALSE, comment=NA}
# Constructing model showing the relationship between BMI and BP
model1 <- lm(BP ~ BMI, data = stats)
model1
summary(model1)

plot(stats$BMI, stats$BP)
abline(model1, col = "yellow")
# Making a report of model
report(model1)
```

The comparison of BMI in the different sampling locations
```{r comparison, echo=FALSE, comment=NA}
# Calculating mean and median
mstat<-stats %>% 
  group_by(Location) %>% 
  summarize(mean = mean(BMI), median = median(BMI))
# Calculating the Variance
vstat<-stats %>% 
  group_by(Location) %>% 
  summarize(variance = var(BMI))
# Merging mstat and vstat dataframes
result<- inner_join(mstat, vstat, by="Location")
result
# Performing a histogram
hist(stats$BMI)
# Making a ggplot of MBI against Location
ggplot(stats, mapping = aes(x = Location, y = BMI)) +labs(title = "ggplot of BMI against Location", x = "Location", y = "BMI") +  # Setting titles
  geom_boxplot() + theme_bw()
# Making summary report of BMI by Location
anov_result <- aov(BMI~Location, stats)
summary(anov_result)
report(anov_result)
```

How disease status depends on both BMI and BP
```{r dependance, echo=FALSE, comment=NA}
# Fitting Generalized Linear Models
dependance <- glm(Status ~ BMI + BP, data = stats, family = "binomial")
summary(dependance)
report(dependance)
```
